---
title: "Assignment 1_Part D_Comprehensive Final Report"
author: "Tran Huu Hao Nguyen"
date: "2025-08-15"
output:
  pdf_document:
    latex_engine: xelatex
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)


# Set default CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))
```


## loading the necessary packages


```{r }
#Load the required packages:
install.packages(c("caret", "rpart", "rpart.plot", "randomForest", "ROCR", "e1071"))
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(ROCR)
library(e1071)

```


## III.	Research Methodology:

## Load the dataset:



```{r }
# Load CSV data
data <- read.csv("enhanced_student_habits_performance_dataset.csv")
data
```


## Explore data structure and summary:


```{r }
# Check structure and summary
str(data)
summary(data)

```

## a)	Phase 1: Data Pre-processing:

## Check for missing values:


```{r }
# Check for any missing values
colSums(is.na(data))
# Expectation: all columns should return 0

```



##  Standardize continuous numeric variables:


```{r }
# Define numeric variables for scaling
num_vars <- c("study_hours_per_day", "screen_time", "sleep_hours",
              "exercise_frequency", "mental_health_rating", "stress_level",
              "exam_anxiety_score", "parental_support_level", "previous_gpa",
              "time_management_score")

# Apply Z-score normalization
data[num_vars] <- lapply(data[num_vars], scale)

```


## Handle outliers with winsorization:

```{r }
# Install DescTools package

install.packages("DescTools")

library(DescTools)


# Custom Winsorization Function
winsorize_manual <- function(x, lower = 0.01, upper = 0.99) {
  qnt <- quantile(x, probs = c(lower, upper), na.rm = TRUE)
  x[x < qnt[1]] <- qnt[1]
  x[x > qnt[2]] <- qnt[2]
  return(x)
}

# Apply to selected columns
winsor_cols <- c("study_hours_per_day", "sleep_hours", "screen_time")

for (col in winsor_cols) {
  data[[col]] <- winsorize_manual(data[[col]])
}

```


## b)	Phase 2: Feature Engineering:

## Create binary target variable for classification:


```{r }
# Create a binary column for pass/fail (1 if exam_score >= 50)
data$pass_exam <- ifelse(data$exam_score >= 50, 1, 0)

```


## Feature transformation and aggregation:


```{r }
# Remove non-predictive identifier
data$student_id <- NULL

# Drop screen-related sub-components since screen_time is already present
data$social_media_hours <- NULL
data$netflix_hours <- NULL


```



## Check: 

```{r }
# Check data structure and summary after preprocessing
str(data)
summary(data)

```


## c)	Phase 3: Model Selection:

## d)	Phase 4: Model Training & Refinement:

## IV.	Experimental Evaluation:



We combine these steps to code them together to give a connection between the content of the research paper and the code here.


The data set is divided into 80% training data and 20% testing data.



```{r }
install.packages(c("caret", "rpart", "rpart.plot", "randomForest", "ROCR", "e1071"))
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(ROCR)
library(e1071)

data <- read.csv("enhanced_student_habits_performance_dataset.csv")

data$pass_exam <- ifelse(data$exam_score >= 50, 1, 0)


# Train-test split (80% training / 20% testing)
set.seed(123)
split <- createDataPartition(data$pass_exam, p = 0.8, list = FALSE)
train <- data[split, ]
test <- data[-split, ]

```




1. Linear regression:



```{r }
lm_model <- lm(exam_score ~ study_hours_per_day + mental_health_rating + exam_anxiety_score + sleep_hours + screen_time,
               data = train)
lm_pred <- predict(lm_model, newdata = test)
lm_rmse <- RMSE(lm_pred, test$exam_score)
lm_r2 <- R2(lm_pred, test$exam_score)

# Display results
cat("🔹 Linear Regression:\n")
cat("   RMSE:", round(lm_rmse, 2), "| R²:", round(lm_r2, 3), "\n\n")
```



2. Logistic regression:


```{r }
# Logistic Regression (predicting pass/fail)
log_model <- glm(pass_exam ~ study_hours_per_day + mental_health_rating + stress_level + exam_anxiety_score + sleep_hours,
                 data = train, family = "binomial")
log_probs <- predict(log_model, newdata = test, type = "response")
log_pred <- ifelse(log_probs > 0.5, 1, 0)
log_conf <- confusionMatrix(factor(log_pred), factor(test$pass_exam))
log_roc <- prediction(log_probs, test$pass_exam)
log_auc <- performance(log_roc, "auc")@y.values[[1]]

# Display results

cat("🔹 Logistic Regression:\n")
print(log_conf)
cat("   AUC:", round(log_auc, 3), "\n\n")


```



3.	Decision tree: 



```{r }
install.packages("rpart.plot")   
install.packages("caret")   
library(caret)              
library(rpart.plot)

# Decision Tree (classification)
tree_model <- rpart(pass_exam ~ study_hours_per_day + mental_health_rating + stress_level + exam_anxiety_score + sleep_hours,
                    data = train, method = "class", cp = 0.01)
tree_pred <- predict(tree_model, newdata = test, type = "class")
tree_conf <- confusionMatrix(tree_pred, factor(test$pass_exam))


install.packages("ROSE")
library(ROSE)
balanced_data <- ovun.sample(pass_exam ~ ., data = train, method = "over", N = 2*nrow(train))$data

tree_model <- rpart(pass_exam ~ ., data = balanced_data, method = "class")
rpart.plot(tree_model)


# Display results

cat("🔹 Decision Tree:\n")
print(tree_conf)
rpart.plot(tree_model)

```




4.	Random forest:




```{r }
install.packages("randomForest")  
library(randomForest)              


# Random Forest (classification)
rf_model <- randomForest(factor(pass_exam) ~ study_hours_per_day + mental_health_rating + stress_level +
                         exam_anxiety_score + sleep_hours + screen_time + parental_support_level,
                         data = train, ntree = 300, mtry = 3, importance = TRUE)
rf_pred <- predict(rf_model, newdata = test)
rf_conf <- confusionMatrix(rf_pred, factor(test$pass_exam))
rf_imp <- importance(rf_model)

# Display results

cat("🔹 Random Forest:\n")
print(rf_conf)
varImpPlot(rf_model)


```

